New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-14 paper/paper.tex
minor changes, fixed font style in chart
refactored bibliography
some minor changes

\documentclass[a4paper]{scrartcl}
%\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{svg}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...
\usepackage{hyperref}
\usepackage[capitalise,nameinlink,noabbrev]{cleveref}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw HyperLogLog estimate and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.

@@

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
\frac{\symNumReg}{\symCount_0}
\right).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
\end{equation}

@@

Consider following integer distribution
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. The expectation of $2^{-\symRegVal}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegVal})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).

@@

Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
\begin{equation}
\symExpectation\left(
\frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\underset{\symNumReg\rightarrow \infty}{=}
\frac{1}{\symExpectation
\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
=
\frac{1}{\symNumReg \symExpectation\left(2^{-\symRegVal}\right)}.
\end{equation}
Together with \eqref{equ:expectation_power_two} we obtain
\begin{equation}
\symPoissonRate
=
\symExpectation\left(
\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\quad
\text{for $\symNumReg\rightarrow \infty$}.
\end{equation}
Therefore, 
\begin{equation}
\label{equ:rawestimator_2}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\end{equation}
is an asymptotically almost unbiased estimator for the Poisson parameter. Its asymptotic relative bias is bounded by $\symEpsPowerSeriesFunc$. This estimator corresponds to the raw estimator \eqref{equ:intermediate_estimate}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:poisson_approximation}).

@@

Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\left(
2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCount_0/\symNumReg)
\right)}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right),
\end{equation}
which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
\subsection{Corrected raw estimation algorithm}
The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively. Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
\label{fig:corrected_raw_avg_exec_time}
\end{figure}

@@

Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear counting algorithm. In this case, the maximum likelihood method under the Poisson model directly leads to the estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}. Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}. Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works very well for the more  general HyperLogLog case.

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCount_0\right)\cdot
\symHelper\left(
\symX\cdot
\left(
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
\right)\right)
-
\left(\symNumReg-\symCount_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCount_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
\log\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCount_0}
{\symCount_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal} + \frac{\symCount_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
for $\symFunc$ can be found, where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$ and
$\symRegVal_\text{max} := \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegVal'_\text{max}}
\log\left(
1+
\frac{\symNumReg-\symCount_0}
{
2^{\symRegVal'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCount_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}}.
\end{equation}
Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}. Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator. If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimator fails for small and large cardinalities.

@@

\subsection{Performance}
In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch. For different cardinalities we loaded the multiplicity vectors of 1000 precalculated and randomly generated HyperLogLog sketches into main memory. The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the new estimation algorithm. For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3GHz. As can be seen, the average calculation time did never exceed 700\,ns.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}

>>>>

\documentclass[a4paper]{scrartcl}
\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{svg}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...
\usepackage{hyperref}
\usepackage[capitalise,nameinlink,noabbrev]{cleveref}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.

@@

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCount_0).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\!\left(1-\symCardinalityOriginalEstimate/2^{32}\right).
\end{equation}

@@

Consider following integer distribution
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. The expectation of $2^{-\symRegVal}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegVal})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).

@@

Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
\begin{equation}
\symExpectation\left(
\frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\underset{\symNumReg\rightarrow \infty}{=}
\frac{1}{\symExpectation
\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
=
\frac{1}{\symNumReg \symExpectation\left(2^{-\symRegVal}\right)}.
\end{equation}
Together with \eqref{equ:expectation_power_two} we obtain
\begin{equation}
\symPoissonRate
=
\symExpectation\left(
\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\quad
\text{for $\symNumReg\rightarrow \infty$}.
\end{equation}
Therefore, 
\begin{equation}
\label{equ:rawestimator_2}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\end{equation}
is an asymptotically almost unbiased estimator for the Poisson parameter. Its asymptotic relative bias is bounded by $\symEpsPowerSeriesFunc$. This estimator corresponds to the raw estimator \eqref{equ:intermediate_estimate}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:poisson_approximation}).

@@

Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\left(
2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCount_0/\symNumReg)
\right)}
=
\frac{\symNumReg\log\!\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\!\left(\symNumReg/\symCount_0\right),
\end{equation}
which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
\subsection{Corrected raw estimation algorithm}
The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively. Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
\label{fig:corrected_raw_avg_exec_time}
\end{figure}

@@

Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear counting algorithm. In this case, the maximum likelihood method under the Poisson model directly leads to the linear counting  estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}. Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}. Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works well for the more general HyperLogLog case.

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCount_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\left(
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
\right)\right)
-
\left(\symNumReg-\symCount_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCount_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
+
\frac{\symCount_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCount_0}
{\symCount_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal} + \frac{\symCount_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
for $\symFunc$ can be found, where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$ and
$\symRegVal_\text{max} := \max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegVal'_\text{max}}
\log\left(
1+
\frac{\symNumReg-\symCount_0}
{
2^{\symRegVal'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCount_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCount_\symRegVal}{2^\symRegVal}}.
\end{equation}
Note, if the HyperLogLog sketch is in the intermediate range, where $\symCount_0=\symCount_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimate \eqref{equ:intermediate_estimate}. Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator. If $\symCount_0>0$ or $\symCount_{\symRegRange+1}>0$ the bounds do no longer follow the harmonic mean and this is also the reason why the raw estimator fails for small and large cardinalities.

@@

\subsection{Performance}
In order to prove that the new algorithm is not only accurate, but also fast, we investigated the average computation time for estimating the cardinality from a given HyperLogLog sketch. For different cardinalities we loaded the multiplicity vectors of 1000 precalculated and randomly generated HyperLogLog sketches into main memory. The average computation time was determined by cycling over these multiplicity vectors and passing them as input to the new estimation algorithm. For each evaluated cardinality value the average execution time was calculated after 100 cycles which corresponds to 100\,000 algorithm executions. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. All these benchmarks where carried out on an Intel Core i5-2500K clocking at 3.3\,GHz. As can be seen, the average calculation time did never exceed 700\,ns.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3\,GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}
